Thermodynamic response functions in a cell fluid model

### 2024-09-12 paper_text.tex
add acknowledgement
updates after feedback

\section{Introduction}
In this work we continue our study of the thermodynamic behavior of a cell fluid model, which was defined in~\cite{KozitskyKozlovskiiDobush2018book,KozitskyKozlovskiiDobush2020}. In works~\cite{KozlovskiiDobush2020,PylyukDobush2020,PylyukEtAlJML2023,PylyukKozlovskiiDobushUJP2023b} this model was used with a modified Morse potential describing the particle interaction. In particular, the equation of state was obtained in~\cite{KozlovskiiDobush2020} in the zero-mode approximation. This equation of state is used in the current work to calculate thermodynamic response functions, namely the isothermal compressibility, the thermal pressure coefficient, and the thermal expansion coefficient.

@@

\section{\label{sec:eos}The equation of state}
\subsection{Pressure as a function of temperature and chemical potential}
The equation of state obtained in~\cite[see Eq.(42)]{KozlovskiiDobush2020} reads
\begin{equation}
	\label{eq:eosMT}
	Pv\beta = E_\mu(M, T) + M \bar \rho_0 + \frac{1}{2} \tilde D(0) \bar \rho_0^2 - \frac{a_4}{24} \bar \rho_0^4.
\end{equation}
The quantities in the left-hand side of the equation are $P$, the pressure; $\beta = (k_{\rm B} T)^{-1}$, the inverse temperature; $k_{\rm B}$, the Boltzmann constant; $T$, the temperature; $v$, cell volume. The quantities in the right-hand side are, in general, functions of the temperature $T$ and the chemical potential $\mu$. Let us present their expressions explicitly.

@@

The parameter $\chi_0$ is used in~\cite{KozlovskiiDobush2020} to single out a contribution in the Fourier transform of the potential that is treated as a reference system defined in the reciprocal space, and is selected as $\chi_0 = 0.07$~\cite[see Eq.(24)]{KozlovskiiDobush2020}.

@@

The quantity $\beta_c$ denotes the critical value of the inverse temperature. In~\cite[see Eq.(31)]{KozlovskiiDobush2020} it was found that
\begin{equation*}
	\varepsilon\beta_c = 0.200, \quad \frac{k_{\rm B} T_c}{\varepsilon} = 4.995.
\end{equation*}
We also use the reduced temperature defined as $T = k_{\rm B}T/\varepsilon$, and thus its critical value $T^*_c = 4.995.$

Since $p$ is independent of temperature, the coefficients $g_n$ are also independent of temperature. The numerical values for other coefficients used in this paper are the same as those in~\cite[see Eqs.(5), (23), and (24)]{KozlovskiiDobush2020}:
\begin{eqnarray}
	\label{params}
	\chi_0 = 0.07, & \quad \gamma = 1.65, \nonumber\\
	n_0 = 1.521, & \quad R_0/\alpha = 2.9544, \nonumber\\
	\alpha^* = 5.0 & \quad p = 1.0.
\end{eqnarray}

The quantity $\tilde D(0)$ entering equations~\eqref{eq:eosMT} and~\eqref{chem_pot} is a function of temperature
\begin{equation}
	\label{def:D0}
	\tilde D(0) = g_2 - \frac{1}{2} \frac{g_3^2}{g_4} - \frac{1}{\beta W(0)}.
\end{equation}
The condition $\tilde{D}(0) = 0$ defines the critical temperature~\cite[see Eq.(31)]{KozlovskiiDobush2020}
\begin{equation}
	k_{\rm B}T_c = \left(g_2 - \frac{1}{2} \frac{g_3^2}{g_4} \right) (B - 1 + \chi_0) \Phi^{(r)}(0).
\end{equation}

@@

At $T>T_c$ the equation of state of a cell fluid model in terms of the density and the temperature has the following form
\begin{equation}\label{eq:eosNT}
	Pv\beta = E_\eta (\rho^*,T) + \bar M \rho_{n} + \frac{\tilde D(0)}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4.
\end{equation}
The quantity $E_\nu (\rho^*, T)$ in \eqref{eq:ro_nT} is the function $E_\mu (M,T)$ \eqref{eq:E_mu} rewritten in terms of density and temperature taking into account the expression \eqref{eq:M_nT}
\begin{eqnarray}\label{eq:E_nu}
	E_\nu (\rho^*, T) & = & - \frac{\ln (2\pi \beta W(0))}{2 N_v}  +  g_0 
	\nonumber\\
	&& - \frac{\beta W(0)}{2} 
	\left(\bar{M} - g_1 + \frac{g_3}{g_4} \tilde{D}(0) + \frac{g_3^3}{6g_4^2} \right)^{2} 
	\nonumber\\
	&& - \frac{g_3}{g_4} {\bar{M}} - \frac{g_3^2}{2 g_4^2}  \tilde D(0) - \frac{1}{24} \frac{g_3^4}{g_4^3}. 
\end{eqnarray}
For the reduced pressure we write
\begin{eqnarray}
	\label{eq:eosPTn_reduced}
	P^* & = & (1 + \tau)T^*_c \bigg[ E_\eta (\rho^*,T) 
	\nonumber\\
	&& + \bar{M} \rho_{n} + \frac{\tilde D(0)}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4\bigg].
\end{eqnarray}

@@

If it is preferable to use the equation of state in the form $P=P(T, \rho^*)$, see~\eqref{eq:eosNT} and~\eqref{eq:eosPTn_reduced}, then the most suitable expressions for the thermal pressure coefficient follow immediately from the definition~\eqref{def:therm_pres_coef}
\begin{equation}
	\beta_V = \left(\frac{\partial P}{\partial T}\right)_\rho
\end{equation}
and
\begin{equation}
	\label{eq:beta_star_n}
	\beta^*_V = \frac{1}{T^*_c} \left(\frac{\partial P^*}{\partial \tau}\right)_{\rho}.
\end{equation}

\begin{figure}[h!]
	\centering \includegraphics[width=0.5\textwidth]{f3a.pdf}
	\vskip-3mm\caption{The reduced thermal pressure coefficient $\beta^*_V$ as a function of the density $\rho^*$ at different values of relative temperature $\tau > 0$~($T > T_c$). The temperature dependence is weak, resulting in multiple isotherms collapsing onto the same line, making them indistinguishable at the scale of the figure.
	}\label{fig3a}
\end{figure}
\begin{figure}[h!]
	\centering \includegraphics[width=0.5\textwidth]{f3b.pdf}
	\vskip-3mm\caption{The reduced thermal pressure coefficient $\beta^*_V$ as a function of the effective chemical potential $M$ at different values of relative temperature $\tau > 0$~($T > T_c$). 
	}\label{fig3b}
\end{figure}
\begin{figure}[h!]
	\centering \includegraphics[width=0.5\textwidth]{f3c.pdf}
	\vskip-3mm\caption{The reduced thermal pressure coefficient $\beta^*_V$ as a function of the relative temperature $\tau$ at different values of the density $\rho^*$. 
	}\label{fig3c}
\end{figure}

Figure~\ref{fig3a} shows the dependency of the pressure coefficient $\beta^*_V$ on the density $\rho^*$ for various values of temperature above the critical one. The dependency of $\beta^*_V$ on the effective chemical potential $M$ is displayed in Fig.~\ref{fig3b}. The dependency of $\beta^*_V$ on the relative temperature $\tau$ is displayed in Fig.~\ref{fig3c} for a few values of density.

@@

@@

For the quantity $\bar{M}$ from Eq.~\eqref{eq:M_nT} one has
\begin{equation}
	\left( \frac{\partial \bar{M}}{\partial \tau} \right)_{\rho^*} =
	\frac{k_{\rm B}T_c \rho_n}{W(0)} \frac{1 - \omega_0}{1 + \tau\omega_0},
\end{equation}
where
\begin{equation*}
	\omega_0 = \frac{\chi_0 + A_{\gamma}}{B - 1 + \chi_0}.
\end{equation*}
For $\tilde{D}(0)$ from~\eqref{def:D0} one has
\begin{equation}
	\frac{\partial \tilde{D}(0)}{\partial \tau} = -\frac{k_{\rm B}T_c}{W(0)} \frac{1 - \omega_0}{1 + \tau \omega_0}.
\end{equation}
While calculating both derivatives, we take into account
\begin{equation}
	\frac{\partial}{\partial \tau} \left[\beta W(0)\right] = -\frac{\beta_c \Phi^{(r)}(0)}{(1+\tau)^2} (B - 1 - A_{\gamma}).
\end{equation}
The derivative of the quantity $E_\nu$ from~\eqref{eq:E_nu} with respect to temperature $\tau$ is
\begin{eqnarray}
	\frac{\partial E_\nu}{\partial \tau} & = &  -\frac{1}{2}\left(\bar{M} - g_1 + \frac{g_3}{g_4} \tilde{D}(0) + \frac{g_3^3}{6g_4^2} \right)^{2} 
	\frac{\partial \beta W(0)}{\partial \tau} 
	\nonumber\\
	&&-\beta W(0)\left(\bar{M} - g_1 + \frac{g_3}{g_4} \tilde{D}(0) + \frac{g_3^3}{6g_4^2} \right)
	\nonumber\\
	&& \times \left(\frac{\partial \bar{M}}{\partial \tau} + \frac{g_3}{g_4}\frac{\partial\tilde{D}(0)}{\partial \tau}\right)
	\nonumber\\
	&& -\frac{g_3}{g_4}\frac{\partial \bar{M}}{\partial \tau} 
	- \frac{g_3^2}{2g_4^2}\frac{\partial \tilde{D}(0)}{\partial \tau}.
\end{eqnarray}
Collecting these all formulas together, we can explicitly calculate $\beta^*_V$ from~\eqref{eq:beta_star_n}
\begin{eqnarray}
	\beta^*_V & = & E_\eta (\rho^*,T) + \bar{M} \rho_{n} + \frac{\tilde D(0)}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4
	\nonumber\\
	&& + (1+\tau) \left(\frac{\partial E_{\nu}}{\partial \tau} + \frac{\partial \bar{M}}{\partial \tau}\rho_{n} + \frac{\rho_n^2}{2}\frac{\partial \tilde{D}(0)}{\partial \tau}\right).
\end{eqnarray}

>>>>

\section{Introduction}
In this work we continue our study of the thermodynamic behavior of a cell fluid model, which was defined in~\cite{KozitskyKozlovskiiDobush2018book,KozitskyKozlovskiiDobush2020}. This model was used with the Morse potential in studies~\cite{KozlovskiiDobush2020,PylyukDobush2020} and with a modified Morse potential in~\cite{PylyukEtAlJML2023,PylyukKozlovskiiDobushUJP2023b} as potentials describing the particle interaction. In particular, the equation of state was obtained in~\cite{KozlovskiiDobush2020} in the zero-mode approximation. This equation of state is used in the current work to calculate thermodynamic response functions, namely the isothermal compressibility, the thermal pressure coefficient, and the thermal expansion coefficient.

@@

\section{\label{sec:eos}The equation of state}
\subsection{Pressure as a function of temperature and chemical potential}
The equation of state obtained in~\cite{KozlovskiiDobush2020} reads
\begin{equation}
	\label{eq:eosMT}
	Pv\beta = E_\mu(M, T) + M \bar \rho_0 + \frac{1}{2} \tilde D(0) \bar \rho_0^2 - \frac{a_4}{24} \bar \rho_0^4.
\end{equation}
The quantities in the left-hand side of the equation are $P$, the pressure; $\beta = (k_{\rm B} T)^{-1}$, the inverse temperature; $k_{\rm B}$, the Boltzmann constant; $T$, the temperature; $v$, cell volume. The quantities in the right-hand side are, in general, functions of the temperature $T$ and the chemical potential $\mu$. Let us present their expressions explicitly.

@@

The parameter $\chi_0$ is used in~\cite{KozlovskiiDobush2020} to single out a contribution in the Fourier transform of the potential that is treated as a reference system defined in the reciprocal space, and is selected as $\chi_0 = 0.07$~\cite{KozlovskiiDobush2020}.

@@

The quantity $\beta_c$ denotes the critical value of the inverse temperature. In~\cite{KozlovskiiDobush2020} it was found that
\begin{equation*}
	\varepsilon\beta_c = 0.200, \quad \frac{k_{\rm B} T_c}{\varepsilon} = 4.995.
\end{equation*}
We also use the reduced temperature defined as $T^* = k_{\rm B}T/\varepsilon$, and thus its critical value $T^*_c = 4.995.$

Since $p$ is independent of temperature, the coefficients $g_n$ are also independent of temperature. The numerical values for other coefficients used in this paper are the same as those in~\cite[eqs.~(5), (23), and~(24)]{KozlovskiiDobush2020}:
\begin{eqnarray}
	\label{params}
	\chi_0 = 0.07, & \quad \gamma = 1.65, \nonumber\\
	n_0 = 1.521, & \quad R_0/\alpha = 2.9544, \nonumber\\
	\alpha^* = 5.0 & \quad p = 1.0.
\end{eqnarray}

The quantity $\tilde D(0)$ entering equations~\eqref{eq:eosMT} and~\eqref{chem_pot} is a function of temperature
\begin{equation}
	\label{def:D0}
	\tilde D(0) = g_2 - \frac{1}{2} \frac{g_3^2}{g_4} - \frac{1}{\beta W(0)}.
\end{equation}
The condition $\tilde{D}(0) = 0$ defines the critical temperature~\cite{KozlovskiiDobush2020}
\begin{equation}
	k_{\rm B}T_c = \left(g_2 - \frac{1}{2} \frac{g_3^2}{g_4} \right) (B - 1 + \chi_0) \Phi^{(r)}(0).
\end{equation}

@@

At $T>T_c$ the equation of state of a cell fluid model in terms of the density and the temperature has the following form
\begin{equation}\label{eq:eosNT}
	Pv\beta = E_\rho (\rho^*,T) + \bar M \rho_{n} + \frac{\tilde D(0)}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4.
\end{equation}
The quantity $E_\rho (\rho^*, T)$ in \eqref{eq:eosNT} is the same as function $E_\mu (M,T)$ \eqref{eq:E_mu} rewritten in terms of density and temperature taking into account the expression \eqref{eq:M_nT}
\begin{eqnarray}\label{eq:E_nu}
	E_\rho (\rho^*, T) & = & - \frac{\ln (2\pi \beta W(0))}{2 N_v}  +  g_0 
	\nonumber\\
	&& - \frac{\beta W(0)}{2} 
	\left(\bar{M} - g_1 + \frac{g_3}{g_4} \tilde{D}(0) + \frac{g_3^3}{6g_4^2} \right)^{2} 
	\nonumber\\
	&& - \frac{g_3}{g_4} {\bar{M}} - \frac{g_3^2}{2 g_4^2}  \tilde D(0) - \frac{1}{24} \frac{g_3^4}{g_4^3}. 
\end{eqnarray}
For the reduced pressure we write
\begin{eqnarray}
	\label{eq:eosPTn_reduced}
	P^* & = & (1 + \tau)T^*_c \bigg[ E_\rho (\rho^*,T) 
	\nonumber\\
	&& + \bar{M} \rho_{n} + \frac{\tilde D(0)}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4\bigg].
\end{eqnarray}

@@

If it is preferable to use the equation of state in the form $P=P(T, \rho^*)$, see~\eqref{eq:eosNT} and~\eqref{eq:eosPTn_reduced}, then the most suitable expressions for the thermal pressure coefficient follow immediately from the definition~\eqref{def:therm_pres_coef}
\begin{equation}
	\beta_V = \left(\frac{\partial P}{\partial T}\right)_\rho
\end{equation}
and
\begin{equation}
	\label{eq:beta_star_n}
	\beta^*_V = \frac{1}{T^*_c} \left(\frac{\partial P^*}{\partial \tau}\right)_{\rho}.
\end{equation}
Explicit calculation for $\beta^*_V$ is presented in Appendix~\ref{sec:app-a}.

\begin{figure}[h!]
	\centering \includegraphics[width=0.5\textwidth]{f3a.pdf}
	\vskip-3mm\caption{The reduced thermal pressure coefficient $\beta^*_V$ as a function of the density $\rho^*$ at different values of relative temperature $\tau > 0$~($T > T_c$). Multiple isotherms collapsing onto the same line, making them indistinguishable at the scale of the figure.
	}\label{fig3a}
\end{figure}
\begin{figure}[h!]
	\centering \includegraphics[width=0.5\textwidth]{f3b.pdf}
	\vskip-3mm\caption{The reduced thermal pressure coefficient $\beta^*_V$ as a function of the effective chemical potential $M$ at different values of relative temperature $\tau > 0$~($T > T_c$). 
	}\label{fig3b}
\end{figure}
\begin{figure}[h!]
	\centering \includegraphics[width=0.5\textwidth]{f3c.pdf}
	\vskip-3mm\caption{The reduced thermal pressure coefficient $\beta^*_V$ as a function of the relative temperature $\tau$ at different values of the density $\rho^*$. 
	}\label{fig3c}
\end{figure}

Figure~\ref{fig3a} shows the dependency of the pressure coefficient $\beta^*_V$ on the density $\rho^*$ for various values of temperature above the critical one. It is interesting to note that the temperature dependence is very weak in this case, and and multiple isotherms collapse onto the same line and are indistinguishable at the scale of the figure. This is not a surprised behavior, as similar one is observed for the thermal pressure coefficient of the Lennard-Jones fluid~\cite{YigzaweSadus2013} as well. The dependency of $\beta^*_V$ on the effective chemical potential $M$ is displayed in Fig.~\ref{fig3b}. The dependency of $\beta^*_V$ on the relative temperature $\tau$ is displayed in Fig.~\ref{fig3c} for a few values of density.

@@

\section{Acknowledgement}

This work was supported by the National Research Foundation of Ukraine under the project No. 2023.03/0201 .

@@

For the quantity $\bar{M}$ from Eq.~\eqref{eq:M_nT} one has
\begin{equation}
	\left( \frac{\partial \bar{M}}{\partial \tau} \right)_{\rho^*} =
	\frac{k_{\rm B}T_c \rho_n}{W(0)} \frac{1 - \omega_0}{1 + \tau\omega_0},
\end{equation}
where
\begin{equation*}
	\omega_0 = \frac{\chi_0 + A_{\gamma}}{B - 1 + \chi_0}.
\end{equation*}
For $\tilde{D}(0)$ from~\eqref{def:D0} one has
\begin{equation}
	\frac{\partial \tilde{D}(0)}{\partial \tau} = -\frac{k_{\rm B}T_c}{W(0)} \frac{1 - \omega_0}{1 + \tau \omega_0}.
\end{equation}
While calculating both derivatives, we take into account
\begin{equation}
	\frac{\partial}{\partial \tau} \left[\beta W(0)\right] = -\frac{\beta_c \Phi^{(r)}(0)}{(1+\tau)^2} (B - 1 - A_{\gamma}).
\end{equation}
The derivative of the quantity $E_\rho$ from~\eqref{eq:E_nu} with respect to temperature $\tau$ is
\begin{eqnarray}
	\frac{\partial E_\rho}{\partial \tau} & = &  -\frac{1}{2}\left(\bar{M} - g_1 + \frac{g_3}{g_4} \tilde{D}(0) + \frac{g_3^3}{6g_4^2} \right)^{2} 
	\frac{\partial \beta W(0)}{\partial \tau} 
	\nonumber\\
	&&-\beta W(0)\left(\bar{M} - g_1 + \frac{g_3}{g_4} \tilde{D}(0) + \frac{g_3^3}{6g_4^2} \right)
	\nonumber\\
	&& \times \left(\frac{\partial \bar{M}}{\partial \tau} + \frac{g_3}{g_4}\frac{\partial\tilde{D}(0)}{\partial \tau}\right)
	\nonumber\\
	&& -\frac{g_3}{g_4}\frac{\partial \bar{M}}{\partial \tau} 
	- \frac{g_3^2}{2g_4^2}\frac{\partial \tilde{D}(0)}{\partial \tau}.
\end{eqnarray}
Collecting these all formulas together, we can explicitly calculate $\beta^*_V$ from~\eqref{eq:beta_star_n}
\begin{eqnarray}
	\beta^*_V & = & E_\rho (\rho^*,T) + \bar{M} \rho_{n} + \frac{\tilde D(0)}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4
	\nonumber\\
	&& + (1+\tau) \left(\frac{\partial E_{\rho}}{\partial \tau} + \frac{\partial \bar{M}}{\partial \tau}\rho_{n} + \frac{\rho_n^2}{2}\frac{\partial \tilde{D}(0)}{\partial \tau}\right).
\end{eqnarray}
